### 2014-10-25 motivation.tex
Most changes to motivation done

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly}.
Assume that all operations in the figure are on the same
\cf{BankAccount} object. Execution contains two concurrent client
sessions, where one of the sessions executes a \cf{withdraw} followed
by a \cf{getBalance} operation. We say that \c{withdraw} precedes
\cf{getBalance} in \emph{session order}, and visualize it as a
directed \cf{so} edge from the former to the later. Edges labelled
\cf{vis} in the figure capture the \emph{visibility} relation between
operations; an operation is said to be \emph{visible} to other
operation if the later can witness the effects of the former. Two
operations that are not related either via \cf{vis} or \cf{so} are
said to be concurrent. The figure shows an execution where concurrent
\cf{withdraw}s in different sessions succeed independently, but a
subsequent \cf{getBalance} operation that witnesses both the
\cf{withdraw}s reports a negative balance.

@@

\name automates this error-prone task by letting the programmer
declare application-level consistency constraints as \emph{contracts},
obviating the need to understand store-level consistency guarantees,
and relieving programmer from having to match application
requirements with appropriate store-level alternatives. The
cornerstone of declarative reasoning in \name is the programming model
built around operation-based CRDTs~\cite{shapiroCRDT}. A replicated
data type in \name is defined as a tagged union type, where each tag
is the name of a unique operation defined on the data type. 
%types in Quelea are defined as reductions over the set of effects.
%The following data type definitions capture the operations and effects
%on the bank account and user index objects.
For example, following is the definition of the \cf{BankAccount}
data type in \name:
\begin{codehaskell}
data BankAccount =  Deposit Float | Withdraw Float 
                 | GetBalance
\end{codehaskell}

In Quelea, an object state is nothing but the set of effects (called
the \emph{context}) on this object. As such, \cf{Deposit} and
\cf{Withdraw} effects simply include the amount deposited or
withdrawn, and \cf{AddUser} effect includes the user name. Similarly,
on the user index table, where the key is user name, the value is
simply an effect (\cf{AddName}) binding some user id at this key.
\cf{GetBalance} and \cf{GetName} correspond to the corresponding
read-only operations. The complete definition of the operations on
bank account and user index objects is given below.

\begin{codehaskell}
-- Get balance takes no arguments and read-only
getBalance :: [BankAcc] {- context -} -> () {- args -}
  -> (() {- ret val -}, Maybe BankAcc {- effect -})
getBalance ctxt _ = (sum [v | Deposit v (- ctxt]
        - sum [v | Withdraw v (- ctxt], Nothing)

-- withdraw returns True on success
withdraw :: [BankAcc] -> Float -> (Bool, Maybe BankAcc)
withdraw c v = if sel1 $ getBalance c () >= v
               then (True, Just $ Withdraw v)
               else (False, Nothing)

deposit :: [BankAcc] -> v -> ((), Maybe BankAcc)
deposit _ v = ((), Just $ Deposit v)

addUser :: [BankAcc] -> UserName -> ((), Maybe BankAcc)
addUser _ n = ((), Just $ AddUser n)

addName :: [UserIdx] -> UserID -> (Bool, Maybe UserIdx)
addName [] uid = (True, Just $ AddName uid)
addName x:_ _ = (False, Nothing)

getName :: [UserIdx] -> ()
        -> (Maybe UserName, Maybe UserIdx)
getName [] _ = (Nothing, Nothing)
getName (AddUser uid:_) _ = (Just uid, Nothing)
\end{codehaskell}

The definitions are a straight forward encoding of the expected behavior.
Quelea programming model provides the programmer complete freedom regarding the
semantics and desired convergence property of the replicated data type, and
indeed, we can encode the well-known CRDTs~\cite{SSS} in a declarative fashion.
Observe that the operation definitions presented here only capture the
convergence properties of the replicated data type, and are not concerned with
the consistency properties, which is expressed through the contract language.

@@

The contract for the strongly consistent \cf{addName} operation is given below:
\begin{smathpar}
\begin{array}{l}
\rsf{addNameCtrt} = \forall (a:\rsf{AddName}). \\
\qquad \sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} \vee \vis{\cureff}{a} \vee a = \cureff
\end{array}
\end{smathpar}

In the above definition, $\cureff$ represents the current effect --- the effect
emitted by the \cf{addName} operation. The contract simply states our
high-level observation; for any effect $a$ which is also an \cf{AddName} effect
on the same object, it must be the case that $a$ is visible to $\cureff$ or
vice verse, or $a$ is $\cureff$. Here, $\sameobjZ$ and $\visZ$ are primitive
relations. This contract ensures that any two \cf{addName} operations will be
seen by each other. The contract for withdraw is similar to the
\cf{addNameCtrt}.

Since the deposit operation does not have any restrictions, its contract is
simply $\true$. Same is the case for \cf{addUser} operation. The contract for
get balance operation is:
\begin{smathpar}
\begin{array}{l}
\rsf{getBalCtrt} = \forall (a:\rsf{Deposit}), (b:\rsf{Withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \vee (\soZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff} \\
\qquad \vee (\soZ \cap \sameobjZ) (b,\cureff) \Rightarrow \vis{b}{\cureff}
\end{array}
\end{smathpar}

If a withdraw $b$ is visible to the get balance operation, then all deposit
operations $a$ visible to withdraw should be visible to the get balance
operation. This prevents negative balance anomalies. The rest of the contract
says that a get balance operation must witness previous deposit and withdraw
operations on the same object in the same session. This prevents missed update
anomalies.

Similar to contracts on operations, Quelea supports contracts on transactions.
The contract on the \cf{getBalanceName} transaction is given below:
\begin{smathpar}
\begin{array}{l}
\rsf{getBalanceNameCtrt} = \forall (a:\rsf{GetName}), (b:\rsf{GetBalance}), \\
\quad (c:\rsf{AddName}), (d:\rsf{AddUser}). ~\trans{a}{b}{c}{d} \wedge \so{a}{b} \\
\qquad \wedge ~\vis{c}{a} \wedge \sameobj{d}{b} \Rightarrow \vis{d}{b}
\end{array}
\end{smathpar}

$\trans{a}{b}{c}{d}$ says that the action pairs $a,b$ and $c,d$ are in the same
transaction, and the two transactions are distinct. The contract says
specifically forbids the anomaly presented in
Figure~\ref{fig:monotonicityAnomaly}. This is the desired semantics for the
\cf{getUserName} transaction.

The programmer simply defines such contracts on the operations and the
transactions. Quelea logically analyzes the contracts, and maps the
corresponding operation to the precise store consistency and isolation level.
Thus, Quelea equips the programmer with a declarative model for reasoning and
expressing eventually consistent programs.

>>>>

Consider the execution shown in fig.~\ref{fig:unsafeWithdrawAnomaly}.
Assume that all operations in the figure are on the same
\cf{BankAccount} object with initial balance as 100. Execution
contains two concurrent client sessions, where one of the sessions
executes a \cf{withdraw} followed by a \cf{getBalance} operation. We
say that \c{withdraw} precedes \cf{getBalance} in \emph{session
order}, and visualize it as a directed \cf{so} edge from the former to
the later. Edges labelled \cf{vis} in the figure capture the
\emph{visibility} relation between operations; an operation is said to
be \emph{visible} to other operation if the later can witness the
effects of the former. Two operations that are not related either via
\cf{vis} or \cf{so} are said to be concurrent. The figure shows an
execution where concurrent \cf{withdraw}s in different sessions
succeed independently, but a subsequent \cf{getBalance} operation that
witnesses both the \cf{withdraw}s reports a negative balance.

@@

\name automates this error-prone task by letting the programmer
declare application-level consistency constraints as \emph{contracts}.
This obviates the need to understand store-level consistency guarantees,
and relieves the programmer from having to match application
requirements with appropriate store-level alternatives. The
cornerstone of declarative reasoning in \name is the programming model
that draws largely on the concept of operation-based
CRDTs~\cite{shapiroCRDT}. The state of a replicated data object in
\name is nothing but the set of effects applied on the object. For
example, to represent a \cf{BankAccount} object in \name, we first
define an sum type of all possible effects on a \cf{BankAccount}
object:
%types in Quelea are defined as reductions over the set of effects.
%The following data type definitions capture the operations and effects
%on the bank account and user index objects.
\begin{codehaskell}
type Amount = Float
data BankAccountEff = Deposit Amount
                    | Withdraw Amount 
                    | GetBalance
\end{codehaskell}
Subsequently, any \cf{BankAccount} object can be simply represented as a set of
effects of type \cf{BankAccountEff} that took effect on the object.
For example, a \cf{BankAccount} object obtained by performing a
deposit of 100 and a withdrawl of 50 on a fresh bank account is
represented as following:
\begin{center}
  \{\cf{Deposit(100), Withdraw(50)}\}
\end{center}
An operation on a replicated data object in \name is simply defined as
a reduction over the set of effects representing the object, and is
expected to produce a new effect along with a return value. For
instance, following are the definitions of operations on
\cf{BankAccount} type in \name (we represent sets as lists for
clarity):
\GK{Better define getBalance in terms of fold. It goes well with our
narrative.}
\begin{codehaskell}
type BankAccount = [BankAccountEff]
getBalance ::  BankAccount {- object -}
  -> () {- args -}
  -> (Amount {- ret val -}, 
      Maybe BankAccountEff {- effect -})
getBalance ctxt _ = (sum [v | Deposit v (- ctxt]
        - sum [v | Withdraw v (- ctxt], Nothing)

withdraw :: BankAccount -> Amount 
  -> (Bool, Maybe BankAccountEff)
withdraw c v = if fst (getBalance c ()) >= v
               then (True, Just (Withdraw v))
               else (False, Nothing)

deposit :: BankAccount -> Amount
  -> ((), Maybe BankAccountEff)
deposit _ v = ((), Just (Deposit v))
\end{codehaskell}

A \cf{getBalance} operation accepts a \cf{BankAccount} object
represented as a set of \cf{BankAccountEff}s, folds over the set
adding deposits and subtracting withdraws, and returns the resultant
amount, along with a new \cf{GetBalance} effect\footnote{wrapped
inside a \cf{Maybe} type for technical reasons}. A \cf{withdraw}
operation first checks if there is sufficient balance in the account.
If the check passes, it returns \cf{True} along with a new
\cf{Withdraw} effect. Otherwise, it returns \cf{False} and produces no
effect. The definition of \cf{deposit} can be similarly understood.

When a bank account operation is executed at one of the replicas in
response to a client request, \cf{BankAccount} object is constructed
using the effects available on the replica, and the definition of the
current operation is applied on the object. Note the set of effects
constituting the input \cf{BankAccount} object are the only effects
that are \emph{visible} to the current operation.  Formally, we say
that a \emph{visibility} relation exists between the input set of
effects and the newly engendered effect. In contrast, a concurrent
operation executing on a different replica can neither witness the
effect of the current operation, not can its effect be visible to the
current operation.

It is worthwhile to note that the above definitions of
\cf{BankAccount} operations are a straightforward encoding of the
expected behavior.  \name programming model provides the programmer
complete freedom regarding the semantics and desired convergence
property of the replicated data type, and indeed, we can encode the
well-known CRDTs~\cite{SSS} in a declarative fashion.  Observe that
the operation definitions presented here only capture the convergence
properties of the replicated data type, and are not concerned with the
consistency properties, which is expressed through the contract
language.

@@

We have previously observed that two \cf{withdraw} operations cannot
be executed concurrently for the integrity constraint on the account
balance to hold. Informally, this means that for any two distinct
\cf{withdraw} operations $a$ and $b$ on a \cf{BankAccount} object,
either the effect of $a$ must be visible to $b$, or the effect of $b$
must be visible to $a$. \name lets us declare this consistency
requirement as a contract over \cf{withdraw}:
\begin{smathpar}
\begin{array}{l}
\rsf{withdrawCtrt}(b) = \forall a. ~a \neq b \conj \rsf{isWithdraw}(a) \conj \\
\qquad \sameobj{a}{b} \Rightarrow \vis{a}{b} \vee \vis{b}{a} 
\end{array}
\end{smathpar}
The above definition of $\rsf{withdrawCtrt}$ is parameterized over the
effect ($b:\cf{BankAccountEff}$) generated by the \cf{withdraw}
operation. The predicate $\rsf{isWithdraw}$ is true for a
\cf{Withdraw} effect, and false for any other effect. $\sameobj{a}{b}$
is true if and only if $a$ and $b$ are effects over the same
\cf{BankAccount} object; we only require \cf{withdraw} operations on
same bank account to be totally ordered. Operations on other bank
accounts can be executed concurrently. Rest of the contract is a
straightforward translation of its informal presentation discussed
earlier.

For the \cf{getBalance}, we construct a contract  precisely to prevent
the anomalies described previously. The contract is shown below:
\begin{smathpar}
\begin{array}{l}
\rsf{getBalCtrt}(c) = \forall (a,b). ~ \rsf{isDeposit}(a) \conj
  \rsf{isWithdraw}(b) \conj \\
\qquad \vis{a}{b} \wedge \vis{b}{c} \Rightarrow \vis{a}{c} \\
\qquad \vee (\soZ \cap \sameobjZ) (a,c) \Rightarrow \vis{a}{c} \\
\qquad \vee (\soZ \cap \sameobjZ) (b,c) \Rightarrow \vis{b}{c}
\end{array}
\end{smathpar}

\GK{It is better to split this contract into two parts.}
If a withdraw $b$ is visible to the get balance operation, then all
deposit operations $a$ visible to withdraw should be visible to the
get balance operation. This prevents negative balance anomalies. The
rest of the contract says that a get balance operation must witness
previous deposit and withdraw operations on the same object in the
same session. This prevents missed update anomalies.

Since the deposit operation does not have any restrictions, its
contract is simply $\true$.

Note that the above contracts for \cf{withdraw} and \cf{getBalance}
express their consistency requirements \emph{independent} of the
semantics of the underlying store. To write such contracts, a
programmer only needs to reason about the semantics of the application
under the replication model professed by \name. The mapping of
application-level consistency requirements to appropriate store-level
guarantees is done automatically behind the screen. This process is
described in the following two sections.
% Similar to contracts on operations, Quelea supports contracts on transactions.
% The contract on the \cf{getBalanceName} transaction is given below:
% \begin{smathpar}
% \begin{array}{l}
% \rsf{getBalanceNameCtrt} = \forall (a:\rsf{GetName}), (b:\rsf{GetBalance}), \\
% \quad (c:\rsf{AddName}), (d:\rsf{AddUser}). ~\trans{a}{b}{c}{d} \wedge \so{a}{b} \\
% \qquad \wedge ~\vis{c}{a} \wedge \sameobj{d}{b} \Rightarrow \vis{d}{b}
% \end{array}
% \end{smathpar}

% $\trans{a}{b}{c}{d}$ says that the action pairs $a,b$ and $c,d$ are in the same
% transaction, and the two transactions are distinct. The contract says
% specifically forbids the anomaly presented in
% Figure~\ref{fig:monotonicityAnomaly}. This is the desired semantics for the
% \cf{getUserName} transaction.

% The programmer simply defines such contracts on the operations and the
% transactions. Quelea logically analyzes the contracts, and maps the
% corresponding operation to the precise store consistency and isolation level.
% Thus, Quelea equips the programmer with a declarative model for reasoning and
% expressing eventually consistent programs.
